Algorithm explanation

### 2013-08-24 article/algo.tex
TeX

\documentclass[a4paper,12pt]{article}

@@

\title{Algorithm explanation}
\author{Yuto Takei \\ The University of Tokyo}

@@

@@

The algorithm assigns a \emph{predicate template} for all predicate
variables in $G$. A template for a predicate variable $P(x_1, \ldots, x_n)$
is a formula $\psi$ constructed from the following definition.
\begin{align*}
\psi ::= & a_1 x_1 + \cdots + a_n x_n + b \leq 0 \mid \\
& \psi \wedge \psi \mid \psi \vee \psi
\end{align*}
A linear expression in the definition above may also be represented in
a equivalent vector form $\mathbf{a} \mathbf{x} + b \leq 0$ for short.
Parameters $\mathbf{a}$ and $b$ are specified by a linear constraint.

@@

For each Horn term vertex $v \in V_T$, we define
$\Delta(v) = (\mathbf{a}_v \mathbf{x}_v \leq b_v, C_v)$
, where
\begin{itemize}
\item $\mathbf{a}_v \mathbf{x}_v \leq b_v$ is a linear inequality that
  contains coefficient variables $\mathbf{a}_v$ and a constant $b_v$,
  and
\item $C_v$ is a set of linear expressions on coefficient variables.
\end{itemize}
Intuitively, it denotes the set of inequalities:
\begin{align*}
\left\lbrace
 \mathbf{a}_v \mathbf{x}_v + b_v \leq 0 \middle|
 \exists \left( \textsc{FV}(C_v)
  \setminus (\mathbf{a}_v \cup \left\lbrace b_v \right\rbrace
 \right); C_v
\right\rbrace
\end{align*}
For a predicate variable vertex, $\mathbf{a}_v \mathbf{x}_v + b_v \leq
0$ serves as a predicate template and $C_v$ describes the parameters
$\mathbf{a}_v$ and $b_v$.

@@

\begin{quote}
If $\models \sigma(C_{v_\bot})$, then $\rho$ defined by
\begin{align*}
 \rho = \left\lbrace
  \left( P, \sigma(\mathbf{a}_v \mathbf{x}_v \leq b_v) \right) \middle|
  \forall v \in V_T; \varphi(v) = P(\mathbf{x}_v)
 \right\rbrace
\end{align*}
is a solution of $G$.
\end{quote}

@@

If a simple solution does not exist for $G$, we try to give a less
simpler solution by transforming $G$ and relaxing the constraint
$C_{v_\bot}$.  This transformation takes place by considering the
unsatisfiable core from the constraint.  Depending on the reason for
the absence of a simple solution, the way of the transformation
differs.

@@

\begin{algorithm}
\caption{$ \solveHorn (G) $}\label{alg:solveHorn}
\begin{algorithmic}
\STATE {$v_\bot \gets $ the root node of $G$}
\STATE {$V_\star \gets \left\lbrace v_\bot \right\rbrace$}
\REPEAT
  \STATE {$C_{v_\bot} \gets \genConstr (G, v_\bot, V_\star)$}
  \IF {$C_{v_\bot}$ is satisfiable}
    \STATE {$\sigma \gets $ one model of $C_{v_\bot}$}
    \STATE {$V_\star \gets V_\star \cup \left\lbrace u \in V_T \mid \forall v \in V_\star; u \leadsto v \right\rbrace$}
  \ELSE
    \STATE {$U \gets $ unsatisfiable core for $C_{v_\bot}$}
    \STATE {$G \gets \transGraph (G, U)$}
    \STATE {$V_\star \gets \left\lbrace v_\bot \right\rbrace$}
  \ENDIF
\UNTIL {$V_\star = V_T$}
% TODO: Consider predicate templates?
\STATE {Build solution by $\sigma$}
\end{algorithmic}
\end{algorithm}
\transGraph procedure determines the unsatisfiable core of the
constraint and transform the Horn graph $G$ and predicate templates.
As mentioned earlier, two ways of graph transformation exist.

\paragraph{Conjunction split}

\paragraph{Disjunction split}

The constraints are first built the weakest so that the disjunctiveness of
whole solution template is fixed first.  The strength matters that the
predicate variables are used as the same expressions depending on
different appearances of the predicate variable.

@@

\subsection{Refinement from Unsatisfiable Core}

An unsatisfiable core $\mathcal{U}$ is a subset of $C_{v_\bot}$.
If a vertex $v \in V$ with the copy origin $U$ satisfies:
\begin{align*}
& \exists t, u \in V; v \leadsto t \wedge v \leadsto u \wedge \\
& \left( \exists i, j, k;
\left\lbrace \left( \mathbf{a}_{s,U,j} = \ldots + \mathbf{a}_{v,U,i} + \ldots \right),
\left( \mathbf{a}_{t,U,k} = \ldots + \mathbf{a}_{v,U,i} + \ldots \right)
\right\rbrace \subseteq \mathcal{U} \right)
\end{align*}
we add $v$ to the copy tree $T$.

>>>>

\documentclass[master,final,11pt]{iscs-thesis}

@@

\etitle{Algorithm explanation}
\eauthor{Yuto Takei}
\jtitle{} \jauthor{}
\esupervisor{}
\jsupervisor{}
\supervisortitle{Professor}
\date{August 28, 2013}

@@

\begin{eabstract}
\end{eabstract}
\begin{jabstract}
\end{jabstract}

@@

The algorithm assigns a \emph{predicate template} for all predicate
variables in $G$. A \emph{predicate templateset} is a mapping from
predicate variables to a predicate template.  A template for a
predicate variable $P(x_1, \ldots, x_n)$ is a formula $\psi$
constructed from the following definition.
\begin{align*}
\psi ::= & a_1 x_1 + \cdots + a_n x_n + b \leq 0 \mid \\
& \psi \wedge \psi \mid \psi \vee \psi
\end{align*}
A linear expression in the definition above may also be represented in
a equivalent vector form $\mathbf{a} \mathbf{x} + b \leq 0$ for short.
Parameters $\mathbf{a}$ and $b$ are specified by a linear constraint.

@@

For each Horn term vertex $v \in V_T$, we define
$\Delta(v) = (\mathbf{a}_v \mathbf{x}_v \leq b_v, C_v)$
, where
\begin{itemize}
\item $\mathbf{a}_v \mathbf{x}_v \leq b_v$ is a linear inequality that
  contains coefficient variables $\mathbf{a}_v$ and a constant $b_v$,
  and
\item $C_v$ is a set of linear expressions on coefficient variables.
\end{itemize}
Intuitively, it denotes the set of inequalities:
\begin{align*}
\left\lbrace
 \mathbf{a}_v \mathbf{x}_v + b_v \leq 0 \mid
 \exists \left( \textsc{FV}(C_v)
  \setminus (\mathbf{a}_v \cup \left\lbrace b_v \right\rbrace
 \right); C_v
\right\rbrace
\end{align*}
For a predicate variable vertex, $\mathbf{a}_v \mathbf{x}_v + b_v \leq
0$ serves as a predicate template and $C_v$ describes the parameters
$\mathbf{a}_v$ and $b_v$.

@@

\begin{quote}
If $\models \sigma(C_{v_\bot})$, then $\rho$ defined by
\begin{align*}
 \rho = \left\lbrace
  \left( P, \sigma(\mathbf{a}_v \mathbf{x}_v \leq b_v) \right) \mid
  \forall v \in V_T; \varphi(v) = P(\mathbf{x}_v)
 \right\rbrace
\end{align*}
is a solution of $G$.
\end{quote}

@@

If a simple solution does not exist for $G$, we try to give a less
simpler solution by transforming $G$ and relaxing the constraint
$C_{v_\bot}$.  This transformation takes place by considering the
unsatisfiable core $\mathcal{U}$.  An unsatisfiable core $\mathcal{U}$
is a subset of $C_{v_\bot}$ which is inconsistent itself.  Depending
on the reason for the absence of a simple solution, the way of the
transformation differs.

@@

\begin{algorithm}
\caption{$ \solveHorn (G) $}\label{alg:solveHorn}
\begin{algorithmic}
\STATE {$v_\bot \gets $ the root node of $G$}
\STATE {$V_\star \gets \left\lbrace v_\bot \right\rbrace$}
\STATE {$T \gets $ initial predicate templateset for $G$}
\REPEAT
  \STATE {$C_{v_\bot} \gets \genConstr (G, v_\bot, V_\star)$}
  \IF {$C_{v_\bot}$ is satisfiable}
    \STATE {$\sigma \gets $ one model of $C_{v_\bot}$}
    \STATE {$V_\star \gets V_\star \cup \left\lbrace u \in V_T \mid \forall v \in V_\star; u \leadsto v \right\rbrace$}
  \ELSE
    \STATE {$\mathcal{U} \gets $ unsatisfiable core for $C_{v_\bot}$}
    \STATE {$(G, T) \gets \transGraph (G, T, \mathcal{U})$}
    \STATE {$V_\star \gets \left\lbrace v_\bot \right\rbrace$}
  \ENDIF
\UNTIL {$V_\star = V_T$}
\RETURN {$\sigma \left( T \right)$}
\end{algorithmic}
\end{algorithm}
\transGraph~procedure determines the unsatisfiable core of the
constraint and transform the Horn graph $G$ and predicate templates.
As mentioned earlier, two ways of graph transformation exist.

\paragraph{Conjunction split}
For the first case previously mentioned, the unsatisfiable core
$\mathcal{U}$ contains coeffcient constraint expressions with a
coefficient parameter of certain predicate variable.  If the predicate
variable $P$ is the cause of unsatisfiability, some constraint
expressions in the unsatisfiable core have a parameter $a_{P,i}$
and/or $b_P$ in the following way.
\begin{align*}
& \exists v_a, v_b \in V_T; v_P \leadsto v_a \wedge v_P \leadsto v_b \wedge \\
& \left( \exists i, j, k;
\left\lbrace \left( \mathbf{a}_{v_a,j} = \ldots + \mathbf{a}_{v_P,i} + \ldots \right),
\left( \mathbf{a}_{v_b,k} = \ldots + \mathbf{a}_{v_P,i} + \ldots \right)
\right\rbrace \subseteq \mathcal{U} \right)
\end{align*}

In such a case, the predicate variable $P$ in question is copied to
$P_1, \ldots, P_n$ so that the different appearances of $P$ on
left-hand of Horn clauses can take different premises.  The solution
template of $P$ becomes the conjunction of $P_1, \ldots, P_n$.

<fig...>

Here, the graph $G=(V,E,\varphi)$ will be transformed into
$G'=(V',E',\varphi')$ such that:
\begin{align*}
V' & = V \setminus \left\lbrace v_P \right\rbrace \cup
  \left\lbrace v_{P_i} \mid \emptyset \right\rbrace \\
E' & = E \\
\varphi' & = \varphi
\end{align*}

Note that after this duplication of the predicate vertex $P$, in order
to induce different linear expressions for $P_1, \ldots, P_n$, it may
be necessary for preceeding predicate variable vertices in $G$ to be
duplicated in the same manner as $P$. It is because that different
derivation of $P$'s implication relation may be required to get
different consequences.

However, succeeding predicate vertices of $P$ in $G$ and other
vertices than predecessors are not affected because those implications
that have $P$ in their assumption can selectively choose the premise
that $P_1, \ldots, P_n$ leads.

Therefore, this graph transformation only affects the predecessors of
$P$.

\paragraph{Disjunction split}
For the second case, the unsatisfiable core $\mathcal{U}$ contains
coefficient constraints that unifies induced linear expressions from
each term in disjunctions to be the same expression.  Similarly as
previous, if the predicate variable $P$ is the cause of
unsatisfiability, some constraint expressions in the unsatisfiable
core have a parameter $a_{P,i}$ and/or $b_P$ in the following way.
\begin{align*}
& \exists v_a, v_b \in V_T; v_P \leadsto v_a \wedge v_P \leadsto v_b \wedge \\
& \left( \exists i, j, k;
\left\lbrace \left( \mathbf{a}_{v_a,j} = \ldots + \mathbf{a}_{v_P,i} + \ldots \right),
\left( \mathbf{a}_{v_b,k} = \ldots + \mathbf{a}_{v_P,i} + \ldots \right)
\right\rbrace \subseteq \mathcal{U} \right)
\end{align*}

In such a case, the predicate variable $P$ in question is separated to
$P_1, \ldots, P_n$ each of which is implied by a term, which was
originally a part of a disjunction to imply $P$.  This intuitively
enables us to avoid generating a common linear expression that
disjunctive terms induces, and to propagate each term's premises to
$P$'s successors instead.

<fig...>

Here, the graph $G=(V,E,\varphi)$ will be transformed into
$G'=(V',E',\varphi')$ such that:
\begin{align*}
V' & = V \setminus \left\lbrace v_P \right\rbrace \cup
  \left\lbrace v_{P_i} \mid \emptyset \right\rbrace \\
E' & = E \\
\varphi' & = \varphi
\end{align*}

After this duplication of the predicate vertex $P$, unlike the
conjunction split, this results in producing a disjunction at
succeeding predicate variables vertices ($R$ in this case).
\begin{align*}
(P_1 \wedge Q) \vee (P_2 \wedge Q) \implies R
\end{align*}
Also, predicate variables that appeared with $P$ in the same
assumption ($Q$ in this case) needs to be duplicated by conjunction,
so as to obtain different premises that go along with $P_i$.
Therefore, the effect of disjunction split may reach to neighboring
predicate vertices.

\paragraph{Combinatorial explosion in disjunctions}
Virtually, if multiple disjunctions exists in an input set of Horn
clauses, a na\"{i}ve way of building a solution is solve all
combinations of choices out of every disjunction and concatenate each
expression with conjunctions and/or disjunctions appropriately.  This
may cause a combinatorial explosion.

Conside example below. $e_i$ represents some conjunctions of linear
expressions.
% TODO: Can't we replace e_i with actual expressions?
\begin{align*}
(e_1 \vee e_2) \implies P \\ (e_3 \vee e_4) \implies Q \\ P \wedge Q \implies \bot
\end{align*}
In this example, a na\"{i}ve algorithm may solve following four
problems separately.
\begin{itemize}
\item $e_1 \implies P_1 \qquad e_3 \implies Q_1 \qquad P_1 \wedge Q_1 \implies \bot$
\item $e_2 \implies P_2 \qquad e_3 \implies Q_2 \qquad P_2 \wedge Q_2 \implies \bot$
\item $e_1 \implies P_3 \qquad e_4 \implies Q_3 \qquad P_3 \wedge Q_3 \implies \bot$
\item $e_2 \implies P_4 \qquad e_4 \implies Q_4 \qquad P_4 \wedge Q_4 \implies \bot$
\end{itemize}
Then the solution is built by follows.
\begin{itemize}
\item $P = ( P_1 \vee P_2 ) \wedge ( P_3 \vee P_4 )$
\item $Q = ( P_1 \vee P_3 ) \wedge ( P_2 \vee P_4 )$
\end{itemize}
However, this method is not efficient.

\paragraph{Algorithm}
To avoid this, our algorithm tries to unify a disjunctions of solution
and determine the necessity of split in conjunctions.

The constraints are first built the weakest so that all conjunctions
are not yet unified.

The strength matters that the
predicate variables are used as the same expressions depending on
different appearances of the predicate variable.

@@
